\myTitle

### 2024-05-15 002-GSI_and_Fst.Rmd
Make zeros blank in assign tables.
Follow up on ERF and RRF hets.

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    #final_baseline = "data/cali-chinook-baseline.rds",
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    ass_table80 = "results/GSI_and_Fst/ass-table-80.pdf",
    rosa_gsi_table = "results/GSI_and_Fst/rosa-gsi-table.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf",
    ass_table80_crop = "results/GSI_and_Fst/ass-table-80-crop.pdf",
    ass_table80_tex = "tex/images/ass-table-80-crop.pdf",
    rosa_gsi_table_crop = "results/GSI_and_Fst/rosa-gsi-table-crop.pdf",
    rosa_gsi_table_tex = "tex/images/rosa-gsi-table-crop.pdf"
  )
}

@@

First we need to join the LFARs on there:
```{r}
# get just the 3 LFAR markers we are using:
lfar_columns <- paste(
  rep(c(
    "NC_037130.1:1062935-1063235",
    "NC_037130.1:828619-828919",
    "NC_037130.1:864908-865208"
  ), each = 2),
  rep(c(1,2), 3),
  sep = "_"
)

@@

### Get some raw numbers

@@

@@

@@

## Identifying fish erroneously included in some collections

The occurrence of winter-run-assigned fish amongst the spring run collections is
a clear indicator that there are some fish in the collections that do not belong
with one another.  The same occurs amongst the fall and the spring run, I expect.

Our goal here is to examine the likelihoods with which fish are assigned to
different reporting units and come up with a common-sense and defensible approach
for discarding individuals that don't belong in the collections they were sampled
from.

We start by tallying up the misassigned fish---those that have a repunit and an
inferred repunit that differ:
```{r}
misassigned_ones <- lfar_top_ass$top %>%
  filter(repunit != inferred_repunit) %>%
  count(repunit, inferred_repunit)

misassigned_ones
```

For each of these 6 pairs, I want to make a scatterplot that shows the repunit
likelihoods of all the fish _from either the repunit or the inferred_repunit_.
The x-axis will be the likelihood for the repunit and the y-axis will be the
likelihood for the inferred_repunit.  So, there is some weird prep to get
this all together.  
```{r}
MO2 <- misassigned_ones %>%
  rename(
    Foc_repunit = repunit,
    Foc_inferred_repunit = inferred_repunit
  )

pick_em <- lapply(1:nrow(MO2), function(i) {

  FR <- MO2$Foc_repunit[i]
  FIR <- MO2$Foc_inferred_repunit[i]

  lfar_top_ass$repu_likes %>%
    ungroup() %>%
    filter(
      (repunit == FR & inferred_repunit == FIR) |
        (inferred_repunit == FR & repunit == FIR)
    ) %>%
    mutate(
      Foc_repunit = FR,
      Foc_inferred_repunit = FIR,
      .before = indiv
    ) 
}) %>%
    bind_rows()
```

OLD STUFF BELOW HERE
```{r}
tmp <- lapply(1:nrow(MO2), function(i) {
  lfar_top_ass$repu_likes %>%
    filter(
      (repunit == MO2$Foc_repunit[i] | repunit == MO2$Foc_inferred_repunit[i]) &
        (inferred_repunit == MO2$Foc_repunit[i] | inferred_repunit == MO2$Foc_inferred_repunit[i])
    ) %>%
    mutate(
      Foc_repunit =  MO2$Foc_repunit[i],
      Foc_inferred_repunit = MO2$Foc_inferred_repunit[i],
      .before = repunit
    )
}) %>%
  bind_rows() %>%
  arrange(indiv, Foc_repunit, Foc_inferred_repunit)

for_plots <- tmp %>%
  mutate(
    rep = ifelse(Foc_repunit == repunit, "rep", "inf"),
    inf = ifelse(Foc_inferred_repunit == inferred_repunit, "inf", "rep")
  ) %>%
  ungroup() %>%
  #count(Foc_repunit, Foc_inferred_repunit, repunit, inferred_repunit, rep, inf)
  select(-repunit, -inferred_repunit) %>%
  pivot_wider(names_from = c(rep, inf), names_sep = "_to_", values_from = repu_sclike) %>%
  mutate(condition = str_c(Foc_repunit, " misassigned to ", Foc_inferred_repunit), .after = Foc_inferred_repunit)
```

Now we should be able to facet wrap those:
```{r}
ggplot(for_plots) +
  geom_point(aes(x = inf_to_rep, y = inf_to_inf, colour = Foc_inferred_repunit)) +
  geom_point(aes(x = rep_to_rep, y = rep_to_inf, colour = Foc_repunit)) +
  facet_wrap(~condition) +
  scale_colour_manual(values = repunit_colors)

So, for each of the six plots, I will need four different repu_sclike's:
For example, for the `repunit = CV-Fall` and `inferred_repunit = CV-Spring` case
we will need to have the repu_sclikes for CV-Fall fish being calculated for CV-Fall
and CV-Spring, and for the CV-Spring fish being calculated for CV-Spring and CV-Fall.

Now, we want to use this to semi-join to get all the likelihoods for these scenarios,
whether or not they were misassigned. 
```{r}
focal_comparisons <- repu_likes %>%
  semi_join(misassigned_ones, by = join_by(repunit, inferred_repunit))
```

And we are going to want to compare these to all the likelihoods of being
from the actual repunit that the fish was col

Now, I want to look at these different comparisons _pairwise_, so we will
just pivot it wider and then look at different columns in there.  This isn't
super "tidy" but I think it should work OK:
```{r}
focal_comparisons %>%
  pivot_wider(names_from = inferred_repunit, values_from = repu_sclike)
```

And I want to look at the observed distribution of these scores in for each
comparison, colored by repunit.
```{r}
ggplot(focal_comparisons, aes(x = ))
```

First thing I am going to do is plot histograms of the,
max likelihoods for all of the ones from each repunit.
```{r}
repunit_order <- unique(pop_labels$repunit)
repu_likes %>%
  mutate(
    repunit_f = factor(repunit, levels = repunit_order),
    inferred_repunit_f = factor(inferred_repunit, levels = repunit_order),
  ) %>%
  ggplot() +
  geom_histogram(aes(x = repu_sclike, fill = repunit_f), colour = "black", linewidth = 0.16) + 
  facet_grid(repunit_f ~ inferred_repunit_f) + 
  scale_fill_manual(values = repunit_colors) +
  theme_bw()
```

```{r}
ta_corr_lfar <- top_ass_w_lfar %>%
  filter(repunit == inferred_repunit)
ta_wrong_lfar <- top_ass_w_lfar %>%
  filter(repunit != inferred_repunit) %>%
  arrange(collection, repu_sclike) %>%
  group_by(collection) %>%
  mutate(y = 190 + (70/9) * 1:n())

ggplot() +
  geom_histogram(
    data = ta_corr_lfar,
    mapping = aes(x = repu_sclike, fill = inferred_repunit)
  ) +
  geom_point(
    data = ta_wrong_lfar,
    mapping = aes(x = repu_sclike, y = y, fill = inferred_repunit),
    size = 3,
    shape = 21
  ) +
  facet_wrap(~coll_f, ncol = 3) +
  scale_fill_manual(values = repunit_colors)
```

Filter out the ones that look like clear mis-labelling and then
count up the mis-assigned fish:
```{r}
tafilt_w_lfar <- top_ass_w_lfar %>%
filter(!(repunit != inferred_repunit & repu_sclike > 0.99))
```

Here we calculate the number of correctly and incorrectly assigned fish:
```{r}
tafilt_w_lfar %>%
  mutate(correct = repunit == inferred_repunit) %>%
  count(correct) %>%
  mutate(fract = n / sum(n))
```

Having the LFAR in there improves things. Good.

Make a table of those:
```{r}

X2_lfar <- tafilt_w_lfar %>%
mutate(
collection_f = factor(collection, levels = collection_order),
inferred_coll_f = factor(inferred_collection, levels = collection_order)
) %>%
count(collection_f, inferred_coll_f, .drop = FALSE) %>%
taf_prep()

Rubias_for_later_lfar <- X2_lfar # keeping this for later...

# get the result
TAF2_lfar <- table_as_figure(
X = X2_lfar,
RC_groups = RC_groups,
external_colors = run_time_colors,
internal_colors = repunit_colors,
plot_margins = c(1, 0.1, 0.7, 0.1)
)

# and plot it
g_lfar <- plot_grid(
TAF2_lfar$full_plot,
plot_grid(
get_legend(TAF2_lfar$for_external_legend),
get_legend(TAF2_lfar$for_internal_legend),
nrow = 1
),
nrow = 2,
rel_heights = c(7,3)
)

g_lfar
```

Adding the LFAR definitely improves the LFAR assignment, though
it still does not get it perfect.
```{r}
ggsave(g_lfar, filename = output_list$assig_table_lfar_all, width = 10, height = 12)
```

@@

```{r}
fbl_no_miss <- full_base2 %>%
  select(-starts_with("NC_037130.1")) %>%  # drop the LFARs
  pivot_longer(
    cols = c(-(indiv:sample_type)),
    names_to = c("locus", "gene_copy"),
    names_pattern = "^(.*)_([12])$"
  ) %>%
  group_by(indiv, locus) %>%
  mutate(isHz = value[1] != value[2]) %>%
  ungroup() %>%
  filter(!is.na(value) & !is.na(isHz))

>>>>

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    ass_table80 = "results/GSI_and_Fst/ass-table-80.pdf",
    rosa_gsi_table = "results/GSI_and_Fst/rosa-gsi-table.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf",
    ass_table80_crop = "results/GSI_and_Fst/ass-table-80-crop.pdf",
    ass_table80_tex = "tex/images/ass-table-80-crop.pdf",
    rosa_gsi_table_crop = "results/GSI_and_Fst/rosa-gsi-table-crop.pdf",
    rosa_gsi_table_tex = "tex/images/rosa-gsi-table-crop.pdf"
  )
}

@@

First we need to join the LFARs on there:
```{r}
# get just the 2 LFAR markers we are using.  We are dropping 828619 because
# it really doesn't do much, and so many reads are off-target.
lfar_columns <- paste(
  rep(c(
    "NC_037130.1:1062935-1063235",
    "NC_037130.1:864908-865208"
  ), each = 2),
  rep(c(1,2), 2),
  sep = "_"
)

@@

## Get some raw numbers

@@

## Make a Plot for scaled likelihood > 0.8

@@

## Make a plot with numbers broken out by the RoSA genotype

@@

## Follow up on those Early alleles found in RRF and ERF

The analysis shows 4 fish from RRF and ERF with heterozygous genoypes (EL).
This has not been previously reported, so we really want to check up on those.
It might be that they had lots of missing data.

So, I want to grab their RoSA genotype strings so that we can verify.
```{r}
# here are the IDs fo the hets from those coastal pops
coastal_hets <- lfar_top_ass$top %>%
  left_join(rosa_genos, by = join_by(indiv)) %>%
  filter(collection %in% c("ERF", "RRF")) %>%
  select(indiv, rosa_geno) %>%
  filter(rosa_geno %in% c("EE", "EL"))
coastal_hets
```

Now, let's get the raw genotype strings (the "hapstrs") off those guys:
```{r}
rosa_hapstrs %>%
  semi_join(coastal_hets, by = join_by(indiv))
```

Those are definitely hets in the left half of the markers.  One of them is also
hets at the right half.  The three remaining ones are L homozygotes on the second
half.  So, there was been some recombination within these.

@@

```{r}
fbl_no_miss <- base_with_lfar %>%
  select(-starts_with("NC_037130.1")) %>%  # drop the LFARs
  pivot_longer(
    cols = c(-(indiv:sample_type)),
    names_to = c("locus", "gene_copy"),
    names_pattern = "^(.*)_([12])$"
  ) %>%
  group_by(indiv, locus) %>%
  mutate(isHz = value[1] != value[2]) %>%
  ungroup() %>%
  filter(!is.na(value) & !is.na(isHz))

### 2024-05-15 tex/body-text-of-paper.tex
Make zeros blank in assign tables.
Follow up on ERF and RRF hets.

The existing SNPtype assays and the additional microhaplotype markers were
developed in the absence of a well-assembled genome for Chinook salmon
\comment{(Is this true?)}. As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both the
existing SNPtype assays and the product of the additional microhaplotype
discovery process we identified the locations of these markers within the genome by mapping the
consensus sequences (roughly 300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002872995.1/}) and version 2  (Otsh\_v2.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_018296145.1/}) genome assemblies,
using both the Burrows-Wheeler Aligner \citep{bwa-mem2009} and the BLAST-like Alignment tool \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers so that we can
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

The population-genetic summary statistics are presented in Table~\ref{tab:pg-summ}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Simple population genetic summaries. All quantities are means
over loci.  $\bar{N}$ is the number of fish with observed genotypes; $\bar{N}_\mathrm{A,ss}$ is the
average number of alleles and $\bar{P}_\mathrm{poly,ss}$ the fraction of polymorphic loci,
after sub-sampling to the smallest sample size per locus; $\bar{H}_\mathrm{exp}$ and
$\bar{H}_\mathrm{obs}$ are expected and observed heterozygosity, respectively. Population
codes are as given in Table~\ref{tab:samples}}
\label{tab:pg-summ}
{\footnotesize
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrrr}
\hline\hline
\input{inputs/popgen-summary.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%
The average number of fish with non-missing data in each population varied from
a low of 39.2 (BCF) to a high of 275.2 (CHLF).  The average number of alleles
per locus, standardized to the smallest sample size, ranged from 2.35 (SRW) to
2.68 (RRF), with those two populations also showing the most extreme values for the
standardized fraction of polymorphic markers at 0.9 and 0.98, respectively.
SRW also had the lowest values for expected and observed heterozygosity:
0.349 and 0.340, respectively. While the highest for expected heterozygosity
was 0.412 (SmRF) and for observed heterozygosity, 0.412 (SmRF).

>>>>

The existing SNPtype assays and the additional microhaplotype markers were
developed prior to the advent of a well-assembled genome for Chinook salmon.
As a consequence, many of these markers had
been used previously without certainty about their location in the genome.
After designing and validating an amplicon-sequencing approach to type both the
existing SNPtype assays and the product of the additional microhaplotype
discovery process we identified the locations of these markers within the genome by mapping the
consensus sequences (roughly 300 bp) around the markers
to both the Chinook salmon version 1 (Otsh\_v1.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_002872995.1/}) and version 2  (Otsh\_v2.0; \url{https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_018296145.1/}) genome assemblies,
using both the Burrows-Wheeler Aligner \citep{bwa-mem2009} and the BLAST-like Alignment tool \citep{kent2002blat}.
A similar mapping exercise was performed for the remaining markers so that we can
report their locations in both Otsh\_v1.0 and Otsh\_v2.0.

@@

The population-genetic summary statistics are presented in Table~\ref{tab:pg-summ}.
%%%%%%%%%%%%%%
\begin{table}
\caption{\footnotesize Simple population genetic summaries. All quantities are means
over loci.  $\bar{N}$ is the number of fish with observed genotypes; $\bar{N}_\mathrm{A,ss}$ is the
average number of alleles and $\bar{P}_\mathrm{poly,ss}$ the fraction of polymorphic loci,
after sub-sampling to the smallest sample size per locus; $\bar{H}_\mathrm{exp}$ and
$\bar{H}_\mathrm{obs}$ are expected and observed heterozygosity, respectively. Population
codes are as given in Table~\ref{tab:samples}}
\label{tab:pg-summ}
{\footnotesize
\begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}} lrrrrr}
\hline\hline
\input{inputs/popgen-summary.tex}
\end{tabular*}
}
\vspace*{-2.3ex}\hrule\vspace*{0.3ex}\hrule
\end{table}
%%%%%%%%%%%%%%%
The average number of fish with non-missing data in each population varied from
a low of 39.2 (BCF) to a high of 275.2 (CHLF).  The average number of alleles
per locus, standardized to the smallest sample size, ranged from 2.35 (SRW) to
2.68 (RRF), with  SRW also having the smallest value of the
standardized fraction of polymorphic markers at 0.9 and RRF sharing the highest value
(0.97) with ERF and BCkF.
SRW also had the lowest values for expected and observed heterozygosity:
0.349 and 0.340, respectively. While the highest for expected heterozygosity
was 0.412 (SmRF) and for observed heterozygosity, 0.411 (CRHS).
